### 2018-07-03 paper/4-build.tex
Wibbles

\vspace{1mm}
We have bolded two different phrases above and, as we will see, the part of the
build system responsible for scheduling tasks in the dependency order
(a `scheduler') can be cleanly separated from the part responsible for deciding
whether a key needs to be rebuilt (a `rebuilder'). We therefore tackle each
aspect separately in subsections~\S\ref{sec-dependency-orderings}
and~\S\ref{sec-out-of-date}.

\subsection{Respecting the dependency order}
\label{sec-dependency-orderings}

Section \S\ref{sec-background} introduced three different \emph{task scehedulers}
that decide which tasks to execute and in what order; see the ``Scheduler'' column
of Table~\ref{tab-summary} in \S\ref{sec-background-summary}.
This subsection explores the properties of the three schedulers, and
possible implementations.

@@

The topological scheduler has two downsides: it is limited to \hs{Applicative}
build systems and requires a fresh topological sort each time. So, while the
actions themselves may be incremental (i.e. unnecessary tasks will not be
performed), the pre-processing is not. We can incrementalise the topological
sort by storing the topological order between build runs, using it as a starting
point in the next run, fixing it up if the build discovers any inconsistency, as
\Excel does.

This approach requires a way to abort tasks that have failed due to out-of-date
dependencies. It is also not minimal in the sense that a task may start, do some
meaningful work, and then abort. However, in the case of an \hs{Applicative}
system, that work is zero.

\Bazel's restarting scheduler does not store the topological order between build
runs; instead, it stores the most recent task dependency information. Since this
information may become outdated, \Bazel may also need to abort a task if a
newly discovered dependency is out-of-date.

@@

% An alternative approach to suspending a task is to abort
% it and restart it again later, at the cost of doing additional work.

\subsection{Determining out-of-date keys} \label{sec-out-of-date}

Suppose the scheduler decides that that a key \hs{k} should be
brought up to date.  The next question is: does any work need to be done,
or is it already up to date?   Or, in a cloud-build system, do we have
a cached copy of the value we need?
These questions can be addressed in one of four
fundamental ways, with a number of tweaks and variations within them.

@@

The idea of a dirty bit is to have one piece of persistent information per key,
saying whether the key is \emph{dirty} or \emph{clean}. After a build, all bits
are set to clean. When the next build starts, anything that changed between the
two states is marked dirty.
% ; and by marking additional things dirty/clean the build system can track
% what needs to rebuild.
If a key and all its transitive dependencies are clean, the key does not need
to be rebuild.

\Excel models the dirty bit approach most directly, having an actual dirty bit
associated with each cell, marking the cell dirty if the user modifies it, and
also making sure all cells that depend on it are recursively marked dirty too.
When rebuilding, if a clean cell only depends on clean cells it is skipped,
otherwise it is rebuilt.

% AM: I didn't understand this bit, so commented it out.
% The only wrinkle in this scheme is that \Excel supports
% monadic tasks, and does not separately record the rebuilt, so it has
% to approximate in this case.

\Make uses file modification times, and compares files to their
dependencies, which can be thought of as a dirty bit which is set when
a file is newer than its dependencies. The interesting property of
this dirty bit is that it is not under the control of \Make; rather it is
existing file-system information that has been repurposed. In particular,
modifying a file automatically clears its dirty bit, and
automatically sets the dirty bit of the nodes depending on it. One
thing \Make does require is that file timestamps only go forward in
time -- something that can be violated by backup software.

When using a dirty bit, it is necessary to check all the dependencies of a key.
For applicative build systems that list is easy to obtain, but for monadic
build systems there is no general way to get all dependencies. Instead \Excel
computes a \emph{static approximation} of the dependencies. For applicative
tasks that approximation is correct. For functions such as \cmd{IF} it marks the
cell dirty if \emph{any} potential dependency has changed, even on the untaken
\cmd{IF} branch. For functions such as \cmd{INDIRECT} whose dependencies cannot
be guessed, it conservatively assumes the dependencies have always changed.

With a dirty bit it is simple to achieve minimality. However, to achieve early
cutoff (\S\ref{sec-background-shake}) it would be important to not set the dirty
bit after a computation that did not change the value. \Excel could use this
approach, but does not. In contrast, \Make cannot implement early cutoff nicely
-- to do so it would have to mark the key clean (so it would not rebuild in the next
run) and at the same time not mark the keys that depend on it dirty -- an
impossible task with only the ability to update to the latest modification time.
\Make can approximate early cutoff by not modifying the result file, and not
marking it clean, but then it will rerun in every subsequent build.

@@

Current build systems using deep constructive traces always record hashes of
terminal \emph{input keys}, but the technique works equally well if we skip any
number $n$ of dependency levels (say $n$ levels). The input-only approach is the
special case of $n = \infty$, and constructive traces are the special case of
$n = 1$.

The deep constructive trace approach only works if the operations are
\emph{deterministic}, otherwise inconsistencies arise
(see~\S\ref{sec-non-determinism}). \simon{That section doesn't actually
  mention ``inconsistencies''.
  Also I think the inconsistencies are called Frankenbuilds, a term we might
  want to introduce.}
Concretely, if the result of building a key
given a set of immediate dependencies can vary, then it is possible that key
\hs{k} will have been computed locally, but something \emph{using} \hs{k} will
have been taken as a constructive input using a different value for \hs{k}. If
another rule combines those two values then two different values of \hs{k} will
have been used in one rule, leading to problems such as link-errors for typical
compilers.

@@

\begin{enumerate}
\item Any system using verifying traces, e.g. \Shake, is unlikely to see
significant benefit from storing more than one \hs{Trace} per key\footnote{There
is a small chance of a benefit if the dependencies change but the result does
not, and then the dependencies change back to what they were before.}.
Therefore, such systems can store \hs{Map}~\hs{k}~\hs{(Trace}~\hs{k}~\hs{v)},
where the initial \hs{k} is the \hs{key} field of \hs{Trace}.
\simon{The data type decl had three args, k, v, r.}

>>>>

\vspace{1mm}
We have emboldened two different aspects above: the part of the
build system responsible for scheduling tasks in the dependency order
(a `scheduler') can be cleanly separated from the part responsible for deciding
whether a key needs to be rebuilt (a `rebuilder'). We tackle each
aspect separately in subsections~\S\ref{sec-dependency-orderings}
and~\S\ref{sec-out-of-date}.

\subsection{The scheduler: respecting the dependency order}
\label{sec-dependency-orderings}

Section \S\ref{sec-background} introduced three different \emph{task schedulers}
that decide which tasks to execute and in what order; see the ``Scheduler'' column
of Table~\ref{tab-summary} in \S\ref{sec-background-summary}.
This subsection explores the properties of the three schedulers, and
possible implementations.

@@

To handle dynamic dependencies we can use the following speculative approach:
build tasks in an arbitrary order, discovering their dependencies on the fly;
whenever a task calls \hs{fetch} on an out-of-date key \hs{dep}, abort the task,
and switch to building the dependency \hs{dep}; eventually the previously
aborted task is restarted and makes further progress thanks to \hs{dep} now
being up to date. When a build completes, we can record the resulting order,
using it as a starting point for the next build and fixing it up if a task's
dependencies have changed, as \Excel does.

This approach requires a way to abort tasks that have failed due to out-of-date
dependencies. It is also not minimal in the sense that a task may start, do some
meaningful work, and then abort.

\Bazel's restarting scheduler does not store the discovered order between build
runs; instead, it stores the most recent task dependency information. Since this
information may become outdated, \Bazel may also need to abort a task if a
newly discovered dependency is out-of-date.

@@

\subsection{The rebuilder: determining out-of-date keys} \label{sec-out-of-date}

Suppose the scheduler decides that that a key should be brought up to date. The
next question is: does any work need to be done, or is it already up to date?
Or, in a cloud-build system, do we have a cached copy of the value we need?
These questions can be addressed in one of four fundamental ways, with a number
of tweaks and variations within them.

@@

The idea of a dirty bit is to have one piece of persistent information per key,
saying whether the key is \emph{dirty} or \emph{clean}. After a build, all bits
are set to clean. When the next build starts, anything that changed between the
two builds is marked dirty. If a key and all its transitive dependencies are
clean, the key does not need to be rebuild.

\Excel models the dirty bit approach most directly, having an actual dirty bit
associated with each cell, marking the cell dirty if the user modifies it, and
also making sure all cells that depend on it are recursively marked dirty too.
To mark dependent cells dirty, it is necessary to statically know task
dependencies. This is possible only for applicative tasks (see \S\ref{sec-deps}).
For monadic tasks, \Excel computes a \emph{static approximation} of task
dependencies: for functions such as \cmd{IF}, the approximation includes all
cells that are mentioned in the formula (the \cmd{IF}'s condition and both
branches), whereas functions such as \cmd{INDIRECT} whose dependencies cannot be
guessed, are conservatively assumed to depend on all cells and are therefore
always marked dirty.

\Make uses file modification times, and compares files to their dependencies,
which can be thought of as a dirty bit which is set when a file is older than
its dependencies. The interesting property of this dirty bit is that it is not
under the control of \Make; rather it is existing file-system information that
has been repurposed. Modifying a file automatically clears its dirty bit, and
automatically sets the dirty bit of the files depending on it (but not
recursively). Note that \Make requires that file timestamps only go forward in
time, which can be violated by backup software.

With a dirty bit it is simple to achieve minimality. However, to achieve early
cutoff (\S\ref{sec-background-shake}) it would be important to reset the dirty
bit after a computation that did not change the value and make sure that cells
that depend on it are not rebuilt unnecessarily. For \Excel, this is difficult
because the dependent cells have already been recursively marked dirty. For
\Make, it is impossible to mark a file clean and at the same time not mark the
keys that depend on it dirty. \Make can approximate early cutoff by not
modifying the result file, and not marking it clean, but then it will be rebuilt
in every subsequent build.

@@

Current build systems using deep constructive traces always record hashes of
terminal \emph{input keys}, but the technique works equally well if we skip any
number of dependency levels (say $n$ levels). The input-only approach is the
special case of $n = \infty$, and constructive traces are the special case of
$n = 1$.

The deep constructive trace approach only works if the operations are
\emph{deterministic}, otherwise it is possible to violate correctness resulting
in a so-called ``Frankenbuild''~\cite{esfahani2016cloudbuild}. Concretely, if
the result of building a key given a set of immediate dependencies can vary,
then it is possible that key \hs{k} will have been computed locally, but
something \emph{using} \hs{k} will have been taken as a constructive input using
a different value for \hs{k}. If another rule combines those two values then two
different values of \hs{k} will have been used in one rule, leading to problems
such as link-errors for typical compilers.

@@

\begin{enumerate}
\item Any system using verifying traces, e.g. \Shake, is unlikely to see
significant benefit from storing more than one \hs{Trace} per key\footnote{There
is a small chance of a benefit if the dependencies change but the result does
not, and then the dependencies change back to what they were before.}.
Therefore, such systems can store \hs{Map}~\hs{k}~\hs{(Trace}~\hs{k}~\hs{v}~\hs{(Hash v))},
where the initial \hs{k} is the \hs{key} field of \hs{Trace}.
